Follow-up Samples

### 2017-02-18 submission/manuscript_outline_20161024.tex
complete Niel edits

\textbf{\emph{Cancer Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were differences
in previously well described carcinoma associated OTUs. These included
the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
stomatis} (Otu001682). There was a decrease in relative abundance in
initial and follow up samples for lesion and carcinoma for
\emph{Parvimonas micra} (P-value \textless{} 0.05), and
\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
contrast, there was no difference in relative abundance in any of these
OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
small percentage of those with adenoma or carcinoma were positive or had
an appreciable relative abundance of any of these respective OTUs
{[}Figure 2{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma. We first investigated this by creating a model to classify
lesion versus normal based on the bacterial community and FIT
measurements. This model had an AUC range of 0.811 - 0.866 after 100
iterations of 20 repeated 10-fold cross validations. The ROC curve for
the final lesion model used was within the observed range of the 100
different iterations {[}Figure 3a{]}. There were a total of 53 variables
{[}Figure S2{]}. For this model FIT measurement resulted in the largest
decrease in MDA while the OTU with the largest MDA was Otu000251
{[}Figure S2a{]}

In order to investigate this we created two different models: one to
classify lesion versus normal and one to classify pre (initial) versus
post (follow up) samples based on the bacterial community and FIT
measurements. T while the initial sample model had an AUC range of 0.485
to 0.686 after 100 iterations of 20 repeated 10-fold cross validations.
By identifying the most important variables for each respective model
and then reducing them to only these factors we were able to increase
the AUC in the lesion model and initial sample model (0.641 - 0.805).\\
The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap with each other {[}Figure 3a{]}. . Interestingly,
the test set AUC range for the initial sample model performed much
better then the training set AUCs. Both the full and reduced initial
sample models overlapped with each other {[}Figure 3b{]} but there was a
marked decrease in the ROC curve for the final before sample model used.

\textbf{\emph{Common OTUs to both Models:}} The reduced models were
built based on the most important variables to either classification
model. whereas for the initial sample model there were a total of 70
variables {[}Figure S3{]}. For both models FIT measurement resulted in
one of the largest decreases in MDA {[}Figure S2a \& S3a{]}. When we
compared the two different reduced models with each other there were a
total of 23 common OTUs. Some of the most common taxonomic
identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
OTUs had classifications to bacteria typically thought of as commensal
{[}Table S2{]}.

\textbf{\emph{Positive Probability after Lesion Removal:}} If there were
common OTUs that could separate adenoma and carcinoma from normal
controls, we would expect to find a decrease in the positive probability
of the follow up sample to be either lesion or an initial sample. This
is what we observed regardless of model used (lesion or initial sample)
or whether it was built on the full or reduced variable data set
{[}Figure 4 \& S4{]} (P-value \textless{} 0.001).

When we separated individuals based on whether they had an adenoma or
carcinoma there was only a decrease in positive probability for the
carcinoma group (P-value \textless{} 0.001) and not for the adenoma
group (P-value \textgreater{} 0.05). We also observed that there were no
significant differences in whether the models classified the samples as
having lesion between the predicted and actual (P-value \textgreater{}
0.05). This lack of difference between the predicted and actual
classifications were also observed for the initial sample model (P-value
\textgreater{} 0.05). Even though the lesion model was not as accurate
in classifying samples as the initial sample model. It was able to
correctly classify the one individual who still had a carcinoma on
follow up {[}Figure 4a \& S4a{]} while the initial sample model did not
{[}Figure 4b \& S4b{]}.

\textbf{\emph{Treatment and Time Differences:}} After observing these
changes in the bacterial community and positive probability we wanted to
assess whether additional treatments, such as chemotherapy and
radiation, could have an impact on the results that we observed. There
was no difference in the amount of change in positive probability for
either the full or reduced lesion model for either chemotherapy (P-value
\textgreater{} 0.05) or radiation therapy (P-value \textgreater{} 0.05).
In contrast, we observed for the the before sample model a significant
difference in decreased positive probability for those treated with
chemotherapy (P-value \textless{} 0.05). All other variables that were
tested showed no difference based on whether chemotherapy or radiation
was used {[}Table S3{]}. Finally, we wanted to know if the length of
time between the initial and follow up sample could be a possible
confounder. Within our study there was a significant difference for the
time elapsed in the collection of the follow up sample between adenoma
and carcinoma (uncorrected P-value \textless{} 0.05), with time passed
being less for adenoma (253 +/- 41.3 days) then carcinoma (351 +/- 102
days).

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group. It should be noted that whether this change was
towards what would be be expected for normal controls is not known.

Although there were no differences when investigating all OTUs, when
looking specifically at four OTUs that taxonomically classified to
previously suggested cancer causing microbes we found that only 2/4 had
a decrease in relative abundance between initial and follow up for those
with carcinoma and 0/4 had differences for those with adenoma. This data
would suggest that these specific OTUs may be important in the
transition of an adenoma to a carcinoma but less so in the initiation of
an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial
microbiome to either be able to classify lesions (adenoma or carcinoma)
or initial samples in order to find common OTUs in the community that
change for both adenoma and carcinoma. What we found was that the
commonly associated CRC bacteria were not highly represented within our
models but rather that OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only the lesion
model having a single OTU from a previously associated cancer bacterium
(\emph{Porphyromonas asaccharolytica}). Using only these important OTUs
and FIT, both models (lesion and initial sample) significantly decreased
positive probability of either lesion or being an initial sample on
follow up {[}Figure 4 \& S4{]}. Further confirmation of the importance
of the changes of commensal bacteria to these classifications was that a
total of 23 OTUs were common to both models and the vast majority
belonged to regular residents of our gut community {[}Table S3{]}.

For the majority of tests performed there were no differences in the
bacterial microbiome based on whether chemotherapy or radiation was
received {[}Table S3{]}. There was a difference in the length of time
between initial and follow up sample between adenoma and carcinoma.
These results would indicate that the findings described were specific
to the surgical intervention and that some of the differences observed
between carcinoma and adenoma samples could be due to differences in
collection time between samples for the two different groups.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool {[}3,4{]} by
exploring what happens to the bacterial community after surgical removal
of a lesion. Based on previous work by Arthur, et al. {[}9{]} it may not
be surprising to have E.coli as one of the most important OTUs and one
that was common to both models. Interestingly, many of the most
important OTUs had taxonomic identification for resident gut microbes.
This could suggest that the bacterial community is one of the first
components that could change during the pathogenesis of disease. These
bacterial microbiome changes could be the first step in allowing more
inflammatory bacterium to gain a foothold within the colon {[}8{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive within our models. There are a number of reasons why this may
have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 4{]}.
It should be noted that our study does not argue against the importance
of these CRC associated bacteria in the pathogenesis of disease but
rather that they are not the main bacteria changing after surgical
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

Many of the common OTUs between the different models used had many OTUs
that taxonomically classified to potential butyrate producers {[}Table
S2{]}. Another batch of OTUs classified to bacteria that can either
degrade polyphenols or are inhibited by them. Both butyrate and
polyphenols are thought to be protective against cancer in part by
reducing inflammation {[}10{]}. These protective compounds are derived
from the breakdown of fiber, fruits, and vegetables by resident gut
microbes. One example of this potential
diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
which was highly prevalent in our models, are known to be increased in
those with high non-meat based protein consumption {[}11{]}. High
protein consumption in general has been linked with an increased CRC
risk {[}12{]}. Conversely, \emph{Bacteroides} are inhibited by
polyphenols which are derived from fruits and vegetables {[}13{]}. Our
data fits with the hypothesis that the microbial metabolites from
breakdown products within our own diet could not only help to shape the
existing community but also have an effect on CRC risk and disease
progression.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}14{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur. The time difference in collection of sample between
adenoma and carcinoma could have affected our observations for
differences between individuals with adenoma or carcinoma. This
confounding though would not affect the observations where these
individuals are grouped together.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}3{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}15{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after surgery and treatment had been completed. This study was approved
by the University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}16{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}17{]}. DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with normal, adenoma, and carcinoma evenly
represented on each run. Each group was randomly assigned to avoid
biases based on sample collection location.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}17{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}18{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals. In brief, the model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and OTUs that had near zero variance were
removed. This pre-processing was performed with the R package caret
(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
samples and making 100 different 80/20 (train/test) splits of the data
where normal and lesion were represented in the same proportion within
both the whole data set and the 80/20 split. Each of these splits were
then run through 20 repeated 10-fold cross validations to optimize the
mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 423 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}19{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate a Benjamini-Hochberg (BH) correction was applied {[}20{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

@@

\hypertarget{ref-arthur_intestinal_2012}{}
9. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
T-J, et al. Intestinal inflammation targets cancer-inducing activity of
the microbiota. Science (New York, N.Y.). 2012;338:120--3.

\hypertarget{ref-louis_gut_2014}{}
10. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
metabolites and colorectal cancer. Nature Reviews Microbiology
{[}Internet{]}. 2014 {[}cited 2017 Feb 14{]};12:661--72. Available from:
\url{http://www.nature.com/doifinder/10.1038/nrmicro3344}

\hypertarget{ref-zhu_intake_2016}{}
11. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
Proteins Substantially Increased the Relative Abundance of Genus
Lactobacillus in Rat Feces. PloS One. 2016;11:e0152678.

\hypertarget{ref-mu_colonic_2016}{}
12. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
Compared with a Normal-Protein Diet. The Journal of Nutrition.
2016;146:474--83.

\hypertarget{ref-ozdal_reciprocal_2016}{}
13. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
Reciprocal Interactions between Polyphenols and Gut Microbiota and
Effects on Bioaccessibility. Nutrients {[}Internet{]}. 2016 {[}cited
2017 Feb 14{]};8:78. Available from:
\url{http://www.mdpi.com/2072-6643/8/2/78}

\hypertarget{ref-cotter_long-term_2016}{}
14. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
et al. Long-Term Follow-up of Patients Having False Positive
Multi-target Stool DNA Tests after Negative Screening Colonoscopy: The
LONG-HAUL Cohort Study. Cancer Epidemiology, Biomarkers \& Prevention: A
Publication of the American Association for Cancer Research, Cosponsored
by the American Society of Preventive Oncology. 2016;

\hypertarget{ref-obrien_impact_2013}{}
15. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.

\hypertarget{ref-daly_evaluation_2013}{}
16. Daly JM, Bay CP, Levy BT. Evaluation of fecal immunochemical tests
for colorectal cancer screening. Journal of Primary Care \& Community
Health. 2013;4:245--50.

\hypertarget{ref-kozich_development_2013}{}
17. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology. 2013;79:5112--20.

>>>>

\textbf{\emph{Carcinoma Associated Bacteria:}} Previous literature has
suggested that a number of oral microbes may be important in CRC
pathogenesis {[}8{]}. So we next examined whether there were differences
in previously well described carcinoma associated OTUs. These included
the OTUs that aligned with \emph{Porphyromonas asaccharolytica}
(Otu000202), \emph{Fusobacterium nucleatum} (Otu000442),
\emph{Parvimonas micra} (Otu001273), and \emph{Peptostreptococcus
stomatis} (Otu001682). There was a decrease in relative abundance in
initial and follow up samples for lesion and carcinoma for
\emph{Parvimonas micra} (P-value \textless{} 0.05), and
\emph{Porphyromonas asaccharolytica} (P-value \textless{} 0.05). In
contrast, there was no difference in relative abundance in any of these
OTUs for those with adenoma {[}Figure 2{]}. We also observed that only a
small percentage of those with adenoma or carcinoma were positive or had
an appreciable relative abundance of any of these respective OTUs
{[}Figure 2{]}.

\textbf{\emph{The Lesion Model:}} We next wanted to identify if there
were any common bacterial microbiome changes in individuals with adenoma
or carcinoma versus normal control individuals. We investigated this by
creating a model to classify lesion versus normal based on the bacterial
community and FIT measurements. This model had an AUC range of 0.811 -
0.866 after 100 iterations of 20 repeated 10-fold cross validations. The
ROC curve for the final lesion model used was within the observed range
of the 100 different test set AUC iterations {[}Figure 3a{]}. There were
a total of 53 variables {[}Figure S2{]}. For this model FIT measurement
resulted in the largest decrease in MDA while the OTU with the largest
MDA was \emph{Ruminococcaceae} (Otu000251) {[}Figure S2a{]}.

If there were common OTUs that could separate adenoma and carcinoma from
normal controls, we would expect to find a decrease in the positive
probability of the follow up sample to be a lesion. This is what we
observed for the lesion model {[}Figure 4{]} (P-value \textless{}
0.001). When we separated individuals based on whether they had an
adenoma or carcinoma there was only a decrease in positive probability
for the carcinoma group (P-value \textless{} 0.001) and not for the
adenoma group (P-value \textgreater{} 0.05). We also observed that there
were no significant differences in whether the models classified the
samples as having lesion between the predicted and actual (P-value
\textgreater{} 0.05). The lesion model was able to correctly classify
the one individual who still had a carcinoma on follow up {[}Figure
4{]}.

\textbf{\emph{The Initial Sample Model:}} After building a model to
classify based on lesion we wanted to then build a separate model
specifically to be able to classify whether samples were initial or
follow up samples based on the bacterial community and FIT measurements.
The initial sample model had an AUC range of 0.485 to 0.686 after 100
iterations of 20 repeated 10-fold cross validations. The test set AUC
range for this model performed better then the training set AUCs. There
was a marked decrease in the ROC curve for the final model used when
compared to the 100 test set AUC iterations {[}Figure 3b{]}. There were
a total of 70 variables {[}Figure S2{]}. For this model
\emph{Ruminococcaceae} (Otu000278) resulted in the largest decrease in
MDA while FIT measurement resulted in the second largest decrease in MDA
{[}Figure S3a{]}.

If there were common OTUs that could separate initial from follow up
sample regardless of whether the lesion was adenoma or carcinoma we
would expect to find a decrease in the positive probability of the
follow up sample to be an initial sample. This is what we observed for
the initial sample model {[}Figure 4{]} (P-value \textless{} 0.001).
When we separated individuals based on whether they had an adenoma or
carcinoma there was a decrease in positive probability for both the
carcinoma group (P-value \textless{} 0.001) and for the adenoma group
(P-value \textless{} 0.001). For this model there was no difference
between the predicted and actual classifications (P-value \textgreater{}
0.05) {[}Figure 4{]}.

\textbf{\emph{Common OTUs to both Models:}} We next wanted to compare
the similarity between the OTU variables used in either model. The main
purpose was to identify which OTUs were important not only for the
classification of lesion but also for the classification of initial or
follow up sample. Potentially, these specific OTUs are the most
important with respect to the bacterial microbiome response to removal
of lesion. When we compared the two different models with each other
there were a total of 23 common OTUs. Some of the most common taxonomic
identifications belonged to \emph{Bacteroides}, \emph{Clostridiales},
\emph{Blautia}, and \emph{Ruminococcaceae}. The vast majority of these
OTUs had classifications to bacteria typically thought of as commensal
{[}Table S2{]}.

\textbf{\emph{Treatment Differences:}} After observing these changes in
the bacterial community and positive probability we wanted to assess
whether additional treatments, such as chemotherapy and radiation, could
have an impact on the results that we observed. There was only a
significant difference for change in positive probability for those
treated with chemotherapy for the initial sample model (P-value
\textless{} 0.05). All other variables that were tested showed no
difference based on whether chemotherapy or radiation was used {[}Table
S3{]}.

@@

From our results there were large observed differences in the bacterial
microbiome between pre and post surgery samples based on whether the
individual had an adenoma or carcinoma. There were much larger
differences between initial and follow up samples based on the thetayc
distance metric and in fecal blood as measured by FIT for individuals
with carcinoma versus adenoma {[}Figure 1{]}. However, there were no
differences between initial and follow up samples for Shannon Diversity,
observed OTUs, or evenness regardless of whether the individual had an
adenoma or carcinoma {[}Table S1{]}. There was also no differences in
relative abundance of any specific OTU for lesion, adenoma only, or
carcinoma only {[}Figure s1{]}. Although there was a detectable change
in the carcinoma group and this change was towards what would be
expected for normal controls, the follow up samples may not be a
completely normal bacterial microbiome.

Although there were no differences when investigating all OTUs, when
looking specifically at four OTUs that taxonomically classified to
previously suggested carcinoma associated microbes we found that only
2/4 had a decrease in relative abundance between initial and follow up
for those with carcinoma and 0/4 had differences for those with adenoma.
This data would suggest that these specific OTUs may be important in the
transition of an adenoma to a carcinoma but less so in the initiation of
an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial
microbiome to be able to classify lesions (adenoma or carcinoma). What
we found was that not only was the positive probability of a lesion
reduced after lesion removal {[}Figure 4 \& S4{]} but also that the
commonly associated CRC bacteria were not highly represented within this
model. Rather, OTUs that made up the most important variables
overwhelmingly belonged to commensal bacteria. With only a single OTU
from a previously associated cancer bacterium (\emph{Porphyromonas
asaccharolytica}). We followed this up by creating a second model to
classify initial versus follow up samples. We found that this model was
able to accurately classify initial versus follow up samples {[}Figure 4
\& S4{]}. Providing additional information on the importance of
commensal bacteria was that there were a total of 23 OTUs in common to
both models and the vast majority belonged to regular residents of our
gut community {[}Table S3{]}.

For the majority of tests performed there were no differences in the
bacterial microbiome based on whether chemotherapy or radiation was
received {[}Table S3{]}. Within our study there was a significant
difference for the time elapsed in the collection of the follow up
sample between adenoma and carcinoma (uncorrected P-value \textless{}
0.05), with time passed being less for adenoma (253 +/- 41.3 days) then
carcinoma (351 +/- 102 days). These results would indicate that the
findings described were specific to the surgical intervention and that
some of the differences observed between carcinoma and adenoma samples
could be due to differences in collection time between samples for the
two different groups. For time differences, specifically, it could
confound the observation that carcinomas changed more than adenomas
{[}Figure 1a \& 1d{]}.

This study builds upon previous work from numerous labs that have looked
into the bacterial microbiome as a potential screening tool
{[}3,4,9--11{]} by exploring what happens to the bacterial community
after surgical removal of a lesion. Based on previous work by Arthur, et
al. {[}12{]} it may not be surprising to have E.coli as one of the most
important OTUs and one that was common to both models. Interestingly,
many of the most important OTUs had taxonomic identification for
resident gut microbes. This could suggest that the bacterial community
is one of the first components that could change during the pathogenesis
of disease. These bacterial microbiome changes could be the first step
in allowing more inflammatory bacterium to gain a foothold within the
colon {[}8{]}.

Curiously, we observed that the typical CRC associated bacteria were not
predictive ewithin our models. There are a number of reasons why this
may have occurred. First, is that they were not present in enough
individuals to be able to classify those with and without disease with a
high degree of accuracy. Second, is that our Random Forest models were
able to gather the same information from measures such as FIT or other
OTUs. It is also possible that both of these explanations could have
played a role. Regardless, our observations would suggest that an
individual's resident bacteria have a large role to play in disease
initiation and could change in a way that allows predictive models to
lower the positive probability of a lesion after surgery {[}Figure 4{]}.
It should be noted that our study does not argue against the importance
of these CRC associated bacteria in the pathogenesis of disease but
rather that they are not the main bacteria changing after surgical
removal of lesion. In fact, it is possible that these CRC associated
bacteria are important in the transition from adenoma to carcinoma and
would be one explanation as to why in our data we not only see high
initial relative abundances in carcinoma and not adenoma individuals but
also large decreases in relative abundance in some of those with
carcinoma but not in those with adenoma after surgery {[}Figure 2{]}.

Many of the common OTUs between the different models used had many OTUs
that taxonomically classified to potential butyrate producers {[}Table
S2{]}. Another batch of OTUs classified to bacteria that can either
degrade polyphenols or are inhibited by them. Both butyrate and
polyphenols are thought to be protective against cancer in part by
reducing inflammation {[}13{]}. These protective compounds are derived
from the breakdown of fiber, fruits, and vegetables by resident gut
microbes. One example of this potential
diet-microbiome-inflammation-polyp axis is that \emph{Bacteroides},
which was highly prevalent in our models, are known to be increased in
those with high non-meat based protein consumption {[}14{]}. High
protein consumption in general has been linked with an increased CRC
risk {[}15{]}. Conversely, \emph{Bacteroides} are inhibited by
polyphenols which are derived from fruits and vegetables {[}16{]}. Our
data fits with the hypothesis that the microbial metabolites from
breakdown products within our own diet could not only help to shape the
existing community but also have an effect on CRC risk and disease
progression.

One limitation of our study is that we do not know whether individuals
who were still classified as positive by the lesion model eventually had
a subsequent CRC diagnosis. This information would help to strengthen
the case for our lesion model keeping a number of individuals above the
cutoff threshold even though at follow up they were diagnosed as no
longer having a lesion. Another limitation is that we do not know if
adding modern tests such as the stool DNA test {[}17{]} could help
improve our overall AUC. This study also drew heavily from those with
Caucasian ancestry making it possible that the observations may not be
representative of those with either Asian or African ancestry. Although
our training and test set are relatively large we still run the risk of
over-fitting or having a model that may not be representative of other
populations. We've done our best to safeguard against this by not only
running 10-fold cross validation but also having over 100 different
80/20 splits to try and mimic the type of variation that might be
expected to occur. The time difference in collection of sample between
adenoma and carcinoma could have affected our observations for
differences between individuals with adenoma or carcinoma. This
confounding though would not affect the observations where these
individuals are grouped together.

@@

\textbf{\emph{Study Design and Patient Sampling:}} The sampling and
design were similar to that reported in Baxter, et al {[}3{]}. In brief,
study exclusion involved those who had already undergone surgery,
radiation, or chemotherapy, had colorectal cancer before a baseline
fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between 1 - 2 weeks after. The bacterial
microbiome has been shown to normalize within this time period {[}18{]}.
Our follow up data set had a total of 67 individuals that not only had a
sample as described but also a follow up sample between 188 - 546 days
after surgery and treatment had been completed. This study was approved
by the University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{FIT and 16S rRNA Gene Sequencing:}} FIT was analyzed as
previously published using both OC FIT-CHEK and OC-Auto Micro 80
automated system (Polymedco Inc.) {[}19{]}. 16S rRNA gene sequencing was
completed as previously described by Kozich, et al. {[}20{]}. DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with normal, adenoma, and carcinoma evenly
represented on each run. Each group was randomly assigned to avoid
biases based on sample collection location.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}20{]}. The general processing workflow
using mothur was as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screened for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The number of sequences for each sample
was rarefied to 10523 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation:}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) with the main testing and training of the model
completed on a data set of 423 individuals. This model was then applied
to our follow up data set of 67 individuals. In brief, the model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and OTUs that had near zero variance were
removed. This pre-processing was performed with the R package caret
(v6.0.73). Optimization of the mtry hyper-parameter involved taking the
samples and making 100 different 80/20 (train/test) splits of the data
where normal and lesion were represented in the same proportion within
both the whole data set and the 80/20 split. Each of these splits were
then run through 20 repeated 10-fold cross validations to optimize the
mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the
Receiver Operator Characteristic). This resulting model was then tested
on the 20\% of the data that was originally held out from this overall
process. Once the ideal mtry was found the entire 423 sample set was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. The default cutoff of 0.5 was used
as the threshold to classify individuals as positive or negative for
lesion. The hyper-parameter, mtry, defines the number of variables to
investigate at each split before a new division of the data is created
with the Random Forest model.

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}22{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between initial and follow up samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate a Benjamini-Hochberg (BH) correction was applied {[}23{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

@@

\hypertarget{ref-yu_metagenomic_2017}{}
9. Yu J, Feng Q, Wong SH, Zhang D, Liang QY, Qin Y, et al. Metagenomic
analysis of faecal microbiome as a tool towards targeted non-invasive
biomarkers for colorectal cancer. Gut. 2017;66:70--8.

\hypertarget{ref-zackular_human_2014}{}
10. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-warren_co-occurrence_2013}{}
11. Warren RL, Freeman DJ, Pleasance S, Watson P, Moore RA, Cochrane K,
et al. Co-occurrence of anaerobic bacteria in colorectal carcinomas.
Microbiome. 2013;1:16.

\hypertarget{ref-arthur_intestinal_2012}{}
12. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
T-J, et al. Intestinal inflammation targets cancer-inducing activity of
the microbiota. Science (New York, N.Y.). 2012;338:120--3.

\hypertarget{ref-louis_gut_2014}{}
13. Louis P, Hold GL, Flint HJ. The gut microbiota, bacterial
metabolites and colorectal cancer. Nature Reviews Microbiology
{[}Internet{]}. 2014 {[}cited 2017 Feb 14{]};12:661--72. Available from:
\url{http://www.nature.com/doifinder/10.1038/nrmicro3344}

\hypertarget{ref-zhu_intake_2016}{}
14. Zhu Y, Lin X, Li H, Li Y, Shi X, Zhao F, et al. Intake of Meat
Proteins Substantially Increased the Relative Abundance of Genus
Lactobacillus in Rat Feces. PloS One. 2016;11:e0152678.

\hypertarget{ref-mu_colonic_2016}{}
15. Mu C, Yang Y, Luo Z, Guan L, Zhu W. The Colonic Microbiome and
Epithelial Transcriptome Are Altered in Rats Fed a High-Protein Diet
Compared with a Normal-Protein Diet. The Journal of Nutrition.
2016;146:474--83.

\hypertarget{ref-ozdal_reciprocal_2016}{}
16. Ozdal T, Sela DA, Xiao J, Boyacioglu D, Chen F, Capanoglu E. The
Reciprocal Interactions between Polyphenols and Gut Microbiota and
Effects on Bioaccessibility. Nutrients {[}Internet{]}. 2016 {[}cited
2017 Feb 14{]};8:78. Available from:
\url{http://www.mdpi.com/2072-6643/8/2/78}

\hypertarget{ref-cotter_long-term_2016}{}
17. Cotter TG, Burger KN, Devens ME, Simonson JA, Lowrie KL, Heigh RI,
et al. Long-Term Follow-up of Patients Having False Positive
Multi-target Stool DNA Tests after Negative Screening Colonoscopy: The
LONG-HAUL Cohort Study. Cancer Epidemiology, Biomarkers \& Prevention: A
Publication of the American Association for Cancer Research, Cosponsored
by the American Society of Preventive Oncology. 2016;

\hypertarget{ref-obrien_impact_2013}{}
18. O'Brien CL, Allison GE, Grimpen F, Pavli P. Impact of colonoscopy
bowel preparation on intestinal microbiota. PloS One. 2013;8:e62815.
